---
title: "Lesson 16: Log-binomial Regression"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#C2352F"
date: "05/19/2025"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Lesson 16: Log-binomial Regression"
    html-math-method: mathjax
    highlight-style: ayu
execute:
  echo: true
  freeze: auto  # re-render only when source changes
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(gt)
library(janitor)
library(rstatix)
library(knitr)
library(gtsummary)
library(moderndive)
library(broom) 
library(here) 
library(ggplot2)
library(ggpubr)
library(logbin)

theme_set(theme_minimal() + 
            theme(text = element_text(size = 28)))

load(here("data", "bc_diagnosis.Rda"))
mean_age = mean(bc$Age)
bc = bc %>% mutate(Late_stage_num = as.numeric(Late_stage_diag)-1, 
                   Age_c = Age - mean(Age))

library(aplore3)
mean_age = mean(glow500$age) %>% round()
glow = glow500 %>% mutate(age_c = age - mean_age)
```

# Learning Objectives

1.  Review Generalized Linear Models and how we can branch to other types of regression.


```{css, echo=FALSE}
.reveal code {
  max-height: 80% !important;
}
```

## Log-binomial Regression

::: columns
::: column
-   [**Outcome type:**]{style="color:#70AD47"} binary, yes or no

 

-   [**Example outcomes:**]{style="color:#5B9BD5"}
    -   Food insecurity
    -   Disease diagnosis for patient
    -   Fracture
:::

::: column
-   [**Population model**]{style="color:#ED7D31"} 

$$ \log(\mu) = \beta_0 + \beta_1 X$$

-   [**Interpretations**]{style="color:#D6295E"}
    -   We have log of probability on the left
    -   So exponential of our coefficients will be **risk ratio**
:::
:::

## Logistic regression vs. log-binomial regression

- Outcome is the same: still a binary variable
  - Still modeling with probability of event: $P(Y=1|X)$
- Logistic = logit = log of odds
  - Aim to get **odds ratios** for interpretation (exponential of coefficients)
  - You *can* convert the odds ratios to risk ratios
- Log-binomial = log of probability
  - Aim to get **risk ratio** or prevalence ratio (exponential of coefficients)

 

- Disadvantage of log-binomial: left hand side ($\log(\pi(X))$) is only a positive value, while right hand side can range from $-\infty$ to $\infty$


## A few classes ago: GLOW Study with interactions

-   **Outcome variable:** any fracture in the first year of follow up (FRACTURE: 0 or 1)

-   **Risk factor/variable of interest:** history of prior fracture (PRIORFRAC: 0 or 1)

-   **Potential confounder or effect modifier:** age (AGE, a continuous variable)

-   Fitted **LOGISTIC REGRESSION** model with interactions: $$\begin{aligned} \text{logit}\left(\widehat\pi(\mathbf{X})\right) & = \widehat\beta_0 + \widehat\beta_1  \cdot I(\text{PF}= \text{``Yes"}) + \widehat\beta_2  \cdot Age + \widehat\beta_3  \cdot I(\text{PF} = \text{``Yes"}) \cdot Age \\  
    \text{logit}\left(\widehat\pi(\mathbf{X})\right) & = -1.376 + 1.002  \cdot I(\text{PF}=\text{``Yes"}) + 0.063  \cdot Age - 0.057  \cdot I(\text{PF}=\text{``Yes"}) \cdot Age
    \end{aligned}$$
    
## Logistic regression: Reporting results of GLOW Study with interactions

-   Remember our main covariate is prior fracture, so we want to focuse on how age changes the relationship between prior fracture and a new fracture!

::: columns
::: column
For individuals 69 years old, the estimated odds of a new fracture for individuals with prior fracture is 2.72 times the estimated odds of a new fracture for individuals with no prior fracture (95% CI: 1.70, 4.35). As seen in @fig-plots-1, the odds ratio of a new fracture when comparing prior fracture status decreases with age, indicating that the effect of prior fractures on new fractures decreases as individuals get older. In @fig-plots-2, it is evident that for both prior fracture statuses, the predict probability of a new fracture increases as age increases. However, the predicted probability of new fracture for those without a prior fracture increases at a higher rate than that of individuals with a prior fracture. Thus, the predicted probabilities of a new fracture converge at age \[insert age here\].
:::

::: column
```{r}
#| label: fig-plots
#| fig-cap: "Plots of odds ratio and predicted probability from fitted interaction model" 
#| fig-subcap:
#|   - "Odds ratio of fracture outcome comparing prior fracture to no prior fracture"
#|   - "Predicted probability of fracture" 
#| fig-height: 9
#| layout-ncol: 2
#| echo: false

load(here("lessons", "12_Assessing_fit/data", "glow_results.Rda"))
ggplot(pred_glow2) +
  geom_hline(yintercept = 1) +
  geom_smooth(method = "loess", aes(x = age, y = OR_YN)) +
  theme(text = element_text(size=35), title = element_text(size=35)) +
  labs(x = "Age (years)", y = "Estimated odds ratio")

frac_pred = predict(glow_m3, prior_age, se.fit = T, type="response")
pred_glow = prior_age %>% mutate(frac_pred = frac_pred$fit, 
                                  age = age_c + mean_age)

ggplot(pred_glow) + #geom_point(aes(x = age, y = frac_pred, color = priorfrac)) +
  geom_smooth(method = "loess", aes(x = age, y = frac_pred, color = priorfrac)) +
  theme(text = element_text(size=35), title = element_text(size=35)) + ylim(0,1) +
  labs(color = "Prior Fracture", x = "Age (years)", y = "Predicted Probability of Fracture")
```
:::
:::


    
## What would a log-binomial regression model look like?

- For the same GLOW data, we could fit a log-binomial model:

$$\text{log}\left(\widehat\pi(\mathbf{X})\right) = \widehat\beta_0 + \widehat\beta_1  \cdot I(\text{PF}= \text{``Yes"}) + \widehat\beta_2  \cdot Age + \widehat\beta_3  \cdot I(\text{PF} = \text{``Yes"}) \cdot Age $$

```{r}
glow_log_binom = glow %>% glm(formula = fracture ~ priorfrac + age_c + priorfrac*age_c, 
                              family = binomial(link = "log"))
```

```{r}
#| echo: false

tidy(glow_log_binom, conf.int = T, exponentiate = T) %>% 
  gt() %>% 
  tab_options(table.font.size = 30) %>%
  fmt_number(decimals = 3)
```

## Issues with log-binomial in `glm`

- What happens when we remove specific people from our study

```{r}
glow2 = glow %>% filter( age < 80 | fracture == "Yes")
```

- Run a logistic regression model using `glm()`

```{r}
glow2_logistic = glow2 %>% 
  glm(formula = fracture ~ priorfrac + age_c + priorfrac*age_c, 
                              family = binomial)
```

- Run a log-binomial model using `glm()`

```{r}
#| eval: false
glow2_log_binom = glow2 %>% 
  glm(formula = fracture ~ priorfrac + age_c + priorfrac*age_c, 
                              family = binomial(link = "log"))
```

`Error: no valid set of coefficients has been found: please supply starting values`

## How to fix this?

- We can fit the log-binomial model using `logbin()`
  - This function uses something called the EM algorithm to fit the data
  - Recall that `glm` uses an iteratively reweighted least squares algorithm
  
```{r}
glow3 = glow2 %>%
  mutate(interaction_PF_age = as.numeric(priorfrac)*age_c)

glow2_log_binom = logbin(formula = fracture ~ priorfrac + age_c + interaction_PF_age, 
         data = glow3)

tidy(glow2_log_binom, conf.int = T, exponentiate = T) %>% 
  gt() %>% 
  tab_options(table.font.size = 30) %>%
  fmt_number(decimals = 3)
```

```{r}
#| echo: false
#| eval: false

tidy(glow2_logistic, conf.int = T) %>% 
  gt() %>% 
  tab_options(table.font.size = 30) %>%
  fmt_number(decimals = 3)

tidy(glow2_log_binom, conf.int = T, exponentiate = T) %>% 
  gt() %>% 
  tab_options(table.font.size = 30) %>%
  fmt_number(decimals = 3)
```


## Testing different variables

- We would go through the same procedure as logistic regression 

 

- Use LRT to test multiple variables OR one categorical variable with multiple levels

 

- Use Wald test to construct confidence intervals
  - Or test one continuous or one binary variable

## Interpretations and visualizations: no interactions

- If we have no interactions in the model: 
```{r}
#| eval: false
glow_main = glow %>% glm(formula = fracture ~ priorfrac + age_c, 
                              family = binomial(link = "log"))
```

- Above did not run

- So need to run it with `logbin`
```{r}
glow_main = logbin(formula = fracture ~ priorfrac + age_c, data = glow)
```

```{r}
tidy(glow_main, conf.int = T, exponentiate = T) %>% 
  gt() %>% 
  tab_options(table.font.size = 30) %>%
  fmt_number(decimals = 3)
```

## Interpretations and visualizations: interactions

## Predictions and visualizations

```{r}
#| fig-align: center
#| fig-width: 7
#| fig-height: 4.5

prior_age = expand_grid(priorfrac = c("No", "Yes"), age_c = (55:90)-69)
frac_pred = predict(glow_log_binom, prior_age, se.fit = T, type="response")
pred_glow = prior_age %>% mutate(frac_pred = frac_pred$fit, 
                                  age = age_c + mean_age)

ggplot(pred_glow) + #geom_point(aes(x = age, y = frac_pred, color = priorfrac)) +
  geom_smooth(method = "loess", aes(x = age, y = frac_pred, color = priorfrac)) +
  theme(text = element_text(size=20), title = element_text(size=16)) + ylim(0,1) +
  labs(color = "Prior Fracture", x = "Age (years)", y = "Predicted Probability of Fracture")
```

## Assessing model fit

